---
title: "Eurobarometer Analysis"
format: html
---

## Overview

This analysis explores patterns in voting intention across European countries using Eurobarometer survey data. We focus on the role of political interest, media use, income, and time. Due to substantial missingness in the dataset, the analysis prioritizes countries with sufficient complete observations for reliable estimation.

---

## 1. Setup

```{r}
#| message: false
#| warning: false
library(MASS) # load first (has conflicts)
library(tidyverse) # load AFTER to override conflicts
library(skimr)
library(haven)
library(broom)
library(modelsummary)
```

## 2. Load the Data

```{r}
#| cache: true
source("data_prep.R")
eb_data <- load_eb_data()
```

## 3. Exploratory Data Analysis

### Inspect the data structure

```{r}
view(eb_data)
```

### Check dataset dimensions

```{r}
dim(eb_data)
```

### Summary statistics

```{r}
skim(eb_data)
```

### Column names

```{r}
names(eb_data)
```

## 4. Data Preparation

Data cleaning and preprocessing are implemented in a reusable script (`data_prep.R`) to ensure consistency across both the analysis and the interactive application.

The key steps include:

- Handling missing and miscoded values (e.g., income values ≥ 96 set to missing)
- Converting labelled survey variables into usable categorical formats
- Standardizing country names (e.g., merging Germany-East/West and UK regions)
- Cleaning Likert-scale variables into consistent ordered categories
- Engineering numeric representations of ordinal variables for modeling

This approach ensures that all downstream analysis is based on a consistent and reproducible data pipeline.

## 8. Religion Trends Over Time

### Aggregate at the country-year level

```{r}
agg <- eb_data |>
  group_by(nation1, year) |>
  summarise(
    relimp_mean = mean(relimp_num, na.rm = TRUE),
    n_obs = sum(!is.na(relimp_num)),
    .groups = "drop"
  )
```

### Check years per country

```{r}
agg |>
  group_by(nation1) |>
  summarise(n_years = n())
```

### Filter to usable data

```{r}
agg_clean <- agg |>
  filter(!is.na(relimp_mean)) |>
  filter(n_obs > 50) |>
  group_by(nation1) |>
  filter(n() > 2) |>
  ungroup()
```

```{r}
agg_clean |>
  ggplot(aes(year, relimp_mean)) +
  geom_line(color = "steelblue", linewidth = 0.9) +
  geom_point(color = "steelblue", size = 1.2, alpha = 0.7) +
  facet_wrap(~nation1, ncol = 3) +
  scale_y_continuous(breaks = seq(1.5, 3, 0.5)) +
  labs(
    x = "Year",
    y = "Avg. Religious Importance",
    title = "Religion in European Countries (1972–2002)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.spacing = unit(0.8, "lines"),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  )
```

### Why Religion Was Not Included in the Model

Although religion was explored descriptively, it was not included in the regression model due to data limitations and modeling priorities. The religion variable (`relimp`) has substantial missingness, which would further reduce the already limited number of complete observations if included.

In addition, the analysis focuses on core predictors of political participation — political interest, media use, and income — which are more directly tied to voting behavior. Religion is therefore used for exploratory trend analysis rather than included in the main regression specification.

## 9. Model Setup

### Identify countries with sufficient complete-case observations

```{r}
eb_data |>
  group_by(nation1) |>
  summarise(
    n = n(),
    valid = sum(!is.na(particip_num) &
      !is.na(income_num) &
      !is.na(polint_num) &
      !is.na(mediause_num)),
    pct_valid = valid / n
  ) |>
  arrange(desc(valid))
```

All models are estimated on complete-case samples to ensure consistent comparison across predictors. While this reduces the sample size, it avoids introducing bias from inconsistent missingness patterns across variables.

### Helper function for regression datasets

```{r}
create_model_data <- function(country_name) {
  eb_data |>
    filter(nation1 == country_name) |>
    filter(
      !is.na(particip_num),
      !is.na(income_num),
      !is.na(polint_num),
      !is.na(mediause_num)
    )
}
```

## 10. Regression Models

### Country Selection

The United Kingdom, France, and the Netherlands were selected based on data availability and comparability. Due to substantial missingness across key variables in the Eurobarometer dataset, only a small subset of observations contains complete information for voting intention, income, political interest, and media use. These three countries provide the largest number of complete cases (approximately 1,600–1,800 observations each), allowing for more reliable model estimation.

In addition, they represent comparable Western European democracies with similar political and media environments, making them suitable for cross-country comparison. This approach prioritizes data quality and consistency over full geographic coverage.

### Model Specification

Income is treated as an ordinal numeric variable to capture overall socioeconomic gradients while simplifying interpretation, rather than modeling each category separately.

```{r}
gb_model <- create_model_data("United Kingdom")
fr_model <- create_model_data("France")
nl_model <- create_model_data("Netherlands")
```

```{r}
c(
  United_Kingdom = nrow(gb_model),
  France = nrow(fr_model),
  Netherlands = nrow(nl_model)
)
```

### OLS models (main analysis)

```{r}
model_gb <- lm(particip_num ~ income_num + polint_num + mediause_num + year, data = gb_model)
model_fr <- lm(particip_num ~ income_num + polint_num + mediause_num + year, data = fr_model)
model_nl <- lm(particip_num ~ income_num + polint_num + mediause_num + year, data = nl_model)

modelsummary(
  list(
    "United Kingdom" = model_gb,
    "France" = model_fr,
    "Netherlands" = model_nl
  ),
  coef_map = c(
    "income_num" = "Income",
    "polint_num" = "Political Interest",
    "mediause_num" = "Media Use",
    "year" = "Year"
  ),
  stars = TRUE,
  gof_map = c("nobs", "r.squared", "adj.r.squared"),
  gof_rename = c(
    nobs = "Observations",
    r.squared = "R²",
    adj.r.squared = "Adj. R²"
  )
)
```

## 11. Robustness Check (ordered logit, UK only)

```{r}
gb_model <- gb_model |>
  mutate(
    particip_ord = factor(
      particip,
      levels = c(
        "Certainly Not",
        "Probably Not",
        "Probably Yes",
        "Certainly Yes"
      ),
      ordered = TRUE
    )
  )
```

```{r}
model_gb_logit <- polr(
  particip_ord ~ income_num + polint_num + mediause_num + year,
  data = gb_model,
  Hess = TRUE
)
summary(model_gb_logit)
```
To account for the ordinal nature of the dependent variable, an ordered logistic regression was estimated for the United Kingdom as a robustness check.

The results are consistent with the OLS specification. Political interest remains the strongest predictor, media use is positively associated with voting intention, and income shows a weak and only marginally significant effect.

While coefficients are expressed in log-odds, the direction and relative importance of predictors remain unchanged. This suggests that the main findings are robust to alternative model specifications.

The magnitude of the political interest coefficient is notably larger than other predictors, reinforcing its dominant role in explaining variation in voting intention.

## 12. Cross-country comparison (OLS only)

```{r}
tidy_gb <- broom::tidy(model_gb) |> mutate(country = "United Kingdom")
tidy_fr <- broom::tidy(model_fr) |> mutate(country = "France")
tidy_nl <- broom::tidy(model_nl) |> mutate(country = "Netherlands")

coef_df <- bind_rows(tidy_gb, tidy_fr, tidy_nl)
```

```{r}
coef_df_filtered <- coef_df |>
  filter(term %in% c("polint_num", "mediause_num", "year")) |>
  mutate(
    term = recode(term,
      polint_num   = "Political Interest",
      mediause_num = "Media Use",
      year         = "Year"
    )
  )
```

```{r}
ggplot(coef_df_filtered, aes(x = term, y = estimate, color = country)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(position = position_dodge(width = 0.5), size = 2.5) +
  geom_errorbar(
    aes(
      ymin = estimate - 1.96 * std.error,
      ymax = estimate + 1.96 * std.error
    ),
    position = position_dodge(width = 0.5),
    width = 0.15,
    linewidth = 0.6
  ) +
  labs(
    x = "",
    y = "Coefficient Estimate",
    title = "Comparison of Key Predictors Across Countries",
    color = "Country"
  ) +
  scale_color_manual(values = c(
    "United Kingdom" = "#4C78A8",
    "France"         = "#F58518",
    "Netherlands"    = "#54A24B"
  )) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey85"),
    plot.title = element_text(face = "bold")
  )
```

## 13. Interpretation (aligned with OLS)

This analysis examines the relationship between likelihood of voting (`particip_num`) and a set of predictors — income, political interest, media use, and time (year) — across three countries: the United Kingdom, France, and the Netherlands. The models are estimated on complete-case samples of 1,771, 1,628, and 1,670 observations, respectively.

Across all three countries, political interest (`polint_num`) is the strongest and most consistent predictor of voting intention. The coefficient is positive and highly statistically significant in each model (≈ 0.16–0.21), indicating that individuals with greater political engagement are substantially more likely to report that they would vote. This relationship is remarkably stable across national contexts.

Media use (`mediause_num`) is also positively associated with voting intention in all three countries, though the magnitude of the effect varies. The relationship is strongest in France (≈ 0.15), moderate in the United Kingdom (≈ 0.11), and somewhat weaker in the Netherlands (≈ 0.09), suggesting that the role of media exposure in shaping political participation may differ across contexts.

Income has a small and inconsistent association with voting intention, suggesting limited explanatory power once political engagement is accounted for. While a modest positive relationship is observed in the United Kingdom and the Netherlands, the effect is not statistically significant in France. Overall, income appears to play a limited role relative to political interest. Results are consistent when modeling income categorically (not shown).

The coefficient for year is negative in both the United Kingdom and France, indicating a slight decline in reported voting likelihood over time, though the effect is weak. In the Netherlands, the time trend is effectively null. Overall, temporal effects appear to be modest and not substantively large.

Model fit is relatively low across all three countries (R² ≈ 0.06–0.08), which is typical for survey-based behavioral data. Notably, the Netherlands exhibits slightly higher explanatory power than the United Kingdom and France, though differences are modest.

Importantly, these models are estimated on subsets of respondents with complete data across all variables. As a result, the findings reflect patterns among respondents who answered all relevant questions, rather than the full sample.

Overall, the results suggest that political engagement — particularly interest in politics — is a robust and cross-national predictor of voting intention, while income plays a limited role. Media exposure also contributes positively, though its influence varies somewhat by country.

These findings suggest that political interest is strongly associated with voting intention, though causal relationships cannot be established with this model. Future work could extend this analysis using multilevel (hierarchical) models to explicitly account for country-level variation.

## 13. Conclusion

Across the United Kingdom, France, and the Netherlands, political interest is the strongest and most consistent predictor of voting intention, with media use also showing a positive but more variable relationship. Income does not appear to play a meaningful role once political engagement is accounted for.

These findings are based on complete-case samples due to substantial missingness, and should be interpreted as descriptive rather than causal. Overall, the results highlight the importance of political engagement in shaping participation.
